A Causal Model of \hbox{DBMS} Suboptimality\\

### 2017-02-23 subopt-tods16/paper_young.tex
characterizing draft

\begin{abstract}
The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest query
execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Infrequently the optimizer
selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
indicated by the existence of a query plan that performs more efficiently
than the DBMS' chosen plan, for the same query. From the engineering
perspective, it is of critical importance to understand the prevalence
of suboptimality and its causal factors.
We study these aspects across \hbox{DBMSes} to identify the
underlying causes. We propose a novel structural causal model to explicate the
relationship between various factors in query optimization and
suboptimality. Our model associates suboptimality with the factors of
complexity of the query and optimizer and concomitant unanticipated
interactions among the components of the optimizer. This model induces a number of specific
hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We observe that suboptimality prevalence
correlates positively with the number of operators available in the \hbox{DBMS}
(one proxy for optimizer complexity) and with the number of plans generated
by the optimizer (another proxy for optimizer complexity) and with
the number of correlation names (a measure of query complexity), providing
empirical support for this model as well as implications for fundamental
improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
limitation of cost-based optimization.\todo{be more specific!}
This paper thus provides
a new methodology to study mature query optimizers, that of empirical
generalization, proposes a novel causal model for query suboptimality, and
tests several hypotheses deriving from this causal model.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]
\terms{}
\keywords{}
\acmformat{\doubleblind{Sabah Currim, Richard T.~Snodgrass, Young-Kyoon Suh, and
 Rui Zhang, 2015, \hbox{DBMS} Metrology: Measuring Query Time.}{A Causal Model of
    \hbox{DBMS} Suboptimality}}

@@

\section{Introduction}
DBMSes underlie all information systems and hence optimizing their performance is of
critical importance. The DBMS's query optimizer plays an important role. But
what if the optimizer {\em doesn't}: what if it selects the wrong plan?

This paper provides a thorough investigation into \hbox{DBMS} {\em suboptimality},
when the \hbox{DBMS} chooses a slower plan over a faster plan for a given query. We systematically examine the factors influencing the number of suboptimal
queries. There could be multiple causes of the suboptimality. One possible
cause could be some peculiarity within the tens of thousands of lines of
code of that query optimizer. Another possible cause could be the query's
complexity. A third possible reason could be some fundamental limitation
{\em within the general architecture of cost-based optimization} that will
always render a good number of queries suboptimal. Prior research in other domains shows that
increasing the complexity negatively influences
performance~\cite{campbell88,moody98}.

To better understand the impact of different factors on suboptimality of
query performance and the interaction between operators, especially in a
dynamic environment, an experimental approach is needed.
Our research introduces a novel approach
to better understand the factors influencing query performance in
\hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
propose an innovative predictive model to understand the presence and influence
of suboptimality in query evaluation. We use an experimental methodology with
the \hbox{DBMS} as a subject to test our hypotheses with respect to factors
influencing suboptimality, applying to 
our experiment data collected over about a \hbox{cumulative} 30~months of query
executions (over 21,000 hours to run almost a million query
executions).\todo{Young: Table I says something different: over 16,000 hours
  ro run almost two million query executions which would correspond to two
  years continuous execution}
Our research falls within creative development of
new evaluation methods and metrics for artifacts, which were identified as
important design-science contributions~\cite{hevner04}.

@@

\item We find that for a surprisingly large portion of queries, the plan
  chosen  by the query optimizer is not the best plan, for some cardinality,
  of the underlying tables.

\item We propose and test a {\em predictive model} for \hbox{DBMSes} to better understand the factors causing suboptimality.

\item \todo{Rick:Mention generational result}

\item The predictive model and these experimental results suggest several specific
  engineering directions.
\end{itemize}

@@

One might ask, shouldn't the task of optimizing queries be left to DBAs?  In
databases, and especially in data warehouses, the number of users writing
and running queries has been growing exponentially. This growth is aided, in
part, by the drag and drop query tools provided by the different
systems. For example, subject areas allow business users to write queries
without knowing anything about the underlying database structure. This,
coupled with constantly growing data presents new challenges for tuning. For
example, \doubleblind{the data warehouse deployed at the University of
  Arizona}{a large data warehouse we are familiar with} runs about
30,000 queries on a daily basis. 
Also, tuning a subject area or tables for one group of queries can
negatively impact the performance of other queries. Query optimization experts often take hours to tune and test existing canned
queries; the amount expended on this one system for manual query optimization
approaches \$100K/year. Therefore, we argue that it is important to
understand how existing query optimizers can be further improved, and
indeed, fundamental limitations inherent in these optimizers.

@@

We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
underlying causes. We have developed an initial predictive causal model that
identifies four 
factors that may play a role in suboptimality. The ultimate goal is to {\em
  understand} a component within a \hbox{DBMS}, its cost-based optimizer, through the
articulation and empirical testing of a general scientific theory.

In Section~\ref{sec:related} we briefly summarize the vast amount of related
work in query optimization to establish the technical basis for our study.
The following section introduces the methodology we will follow, that of
empirical generalization. We then present a predictive, causal model of
suboptimality and empirically test eight specific hypotheses derived from that
model\c2j{. These are the first scientific results that we are aware of that
  test a general causal model of \hbox{DBMSes}.}{ on three \hbox{DBMSes}.
  These are the first scientific results that we are aware of that apply
  {\em across} \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm.} This model has implications for research in
engineering more efficient \hbox{DBMSes}.

@@

The upper graph in Figure~\ref{fig:query769} represents the plans chosen by
a common DBMS as the cardinality of {\tt FT\_HT1} decreases from 2M tuples
to 10K tuples in units of 10K tuples. The \hbox{x-axis} depicts the estimated
cardinality and the y-axis  a plan chosen for an interval of
cardinalities. So Plan~P0 was chosen for 2M tuples, switching to Plan~P1 at
1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through the
sequence P0, P1, P0, P1, P2, P0, P2, and finally P3 at 40,000 tuples.

@@

We have found through our experiments that flutter and suboptimality are all
around us: {\em every} \hbox{DBMS} that we have examined, including several
open source \hbox{DBMSes} and several proprietary
\hbox{DBMSes}, covering much of the installed base worldwide, exhibit these
phenomena, even when optimizing very simple queries. In the Confirmatory
Experiment described in detail in Section~\ref{sec:experiments}, we started
with 6,966 query instances (a query run on a specific \hbox{DBMS}) after our extensive query measurement
protocol~\doubleblind{\todo{ref SIGMOD'13 paper!}}{}. While about 20\%,
1,491, of these query instances contained only one
plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
cardinality in increments of 10K tuples, a total of 200 cardinalities): see
Figure~\ref{fig:planchanges}. Slightly over half, 4,093, exhibited
suboptimality somewhere along those 200 cardinalities; a few had
many changes to a plan that was in fact suboptimal, as indicated in
Figure~\ref{fig:suboptplanchanges}, across all four \hbox{DBMSes} considered.

\begin{figure}[h!]\centering
\includegraphics[width=0.60\textwidth]{figures/org_ncpq.pdf}
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

\begin{figure}[t]\centering
\includegraphics[width=0.60\textwidth]{figures/zi_ncpq.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  number of changes to a {\em sub-optimal}
  plan\label{fig:suboptplanchanges}}
\vspace*{-2ex}
\end{figure}

One oft-stated observation is that the role of query optimization is not to
get the {\em best} plan, but rather to get a plan that is acceptably
good. Figure~\ref{fig:suboptcumulative} shows
the cumulative distribution of the relative amount of suboptimality (where
an $x$-value of 100 denotes that the query ran 100\% slower than the optimal
query, that is, twice as long). The good news is that 2,738 query instances,
or 67\% of the suboptimal queries, exhibited only a small degree of
suboptimality: less than 30\%. The challenge is that over fifth of all
queries (1,355) exhibited a significant amount of suboptimality ($>= 30\%$); the
relative slowdown can be quite large for some queries.

We started with 7,640 query instances (a particular query running on a
particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
in Section~\ref{sec:experiments}. After our protocol, we were left with
  6,966 query instances, of which 5,475 had at least one change point, so those are
  the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
4,093 (75\%) had some suboptimality.

@@

We emphasize three important points. First, we used a sophisticated query
measurement methodology that reduces the measurement variance, so that the
query plans we identify as suboptimal definitely are so. Second, these
results are over four \hbox{DBMSes}, and thus, such phenomena are not dependent on
a particular implementation of cost-based optimization. Rather, they seem to
be common to {\em any} cost-based optimizer, independent of the specific
cardinality \hbox{estimation} or plan costing or plan enumeration algorithm or
implementation. Third, suboptimal plans are {\em not} the result of poor
coding or of inadequate algorithms. We view query optimization in modern
\hbox{DBMSes} as an engineering marvel, especially given the complexity of the SQL
language and the requirements and expectations of \hbox{DBMS} users, who often
demand that important queries simply not get slower with a new release of
the \hbox{DBMS}. Rather, the prevalence of suboptimality observed here is a
reflection of the complexity of the task of query optimization.

%\pagebreak
This is where the shift to a new methodology comes in. We want to understand
cost-based optimization deeply, in this case to determine if a fundamental
limitation exists, which means that we need to study multiple instances of
that optimization architecture, to achieve generalizable results. To do so,
we will articulate a predictive causal model for how
and in what circumstances suboptimality arises and will provide compelling
evidence via \hbox{hypothesis} testing that this model characterizes the
\hbox{behavior} of query optimizers in general. This is what is meant by ``empirical
generalization'' and why it is needed to answer such questions. In
  the long history of research in database query optimization, or even of
  databases in general, our model and its hypothesis tests are the first
  predictive results that we are aware of that apply {\em across}
  \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm. This is due to the adoption of a new research
  methodology, that of empirical generalization. Such a causal model can
\hbox{provide} guidance to the community about where fundamental research is needed
and to the \hbox{DBMS} engineers about where to focus their efforts.

@@

SQL has emerged as the {\em de facto} and {\em de jure} standard relational
database query and update language. It emerged as an ANSI and ISO standard
in 1986--7, with a major extension and refinement as SQL-92~\cite{Melton93} 
and an even larger extension as SQL:1999~\cite{Melton03} and refinement as
SQL:2008~\cite{SQL2008} and SQL:2011.

There has been extensive work in query optimization over the last 40
years~\cite{Ioannidis96,Jarke84}. Query optimization and evaluation
proceeds in several general steps~\cite{Ramakrishnan03}. First, the SQL query is 
translated into alternative query evaluation plans based on the 
relational algebra via {\em query enumeration}. The cost
of each plan is then estimated and the plan with the lowest estimated cost
is chosen. These steps comprise query optimization, specifically {\em
cost-based query optimization}~\cite{Selinger}. The selected query plan is
then evaluated by the query execution engine which implements a set of
physical operators, often several for each logical operator such as
join~\cite{Graefe93}.

@@

An optimizer for a language like SQL must contend with a huge search space of
complex queries. Its first objective must be {\em correctness}: that the
resulting query evaluation plan produce the correct result for the
query. A secondary but clearly very important objective is {\em efficiency}; after
all, that is the raison d'etre for this phase. As is well known, the name
for this phase is an exaggeration, as existing optimizers do not produce
provably optimal plans. That said, the
query optimizers of prominent \hbox{DBMSes} generally do a superb job of producing
the best query evaluation plan for most queries. This performance is the
result of a fruitful collaboration between the research community and
developers.

@@

However, even with great effort over decades, optimizers as a general class
are still poorly understood. As has been observed, ``query optimization has
acquired the dubious reputation of being something of a black
art''~\cite{Babcock05}. DeWitt has gone farther, stating that ``query
optimizers [do] a terrible job of producing reliable, good plans [for
  complex queries] without a lot of hand
tuning''~\cite[page~59]{winslett02}. And as we will see, suboptimality is
possible even when considering only simple queries.

While this paper does not provide direct solutions to address
suboptimality, we envision that by adopting our proposed predictive model
for suboptimality, engineering practice, such as dynamic reoptimization just
mentioned may benefit. We elaborate on this subject in
Sections~\ref{sec:root} and \ref{sec:engineering}, where we discuss the engineering implications
of our causal model.

@@

For the construct of {\em query complexity} we have identified four
variables: ``Number of correlation names'', ``Presence of aggregate'',
``Presence of primary key attribute'', and ``Presence of subquery". For each independent
variable, we have interventional control in our experiments, in that we can
manipulate the values of these variables through the construction of the
actual query to be optimized. The first is the number of correlation names
defined in the FROM clauses. The second is whether a single aggregate operator
(that is, either {\tt Max} or {\tt Avg}) appears in the SELECT clause.
The third
is whether a primary key attribute appears in at least one predicate in the
WHERE clause. The fourth is whether a single subquery appears in the
WHERE clause; that subquery evaluates to a single value that is
equality-compared with an attribute, in the where clause. The rationale is
that each factor may expand the number of plans or otherwise render the search for
the optimal plan more complex.

@@

In the middle of the figure is the construct of {\em plan space
complexity}. Given a particular query, its measurable
complexity will impact the total number of candidate plans considered
by the optimizer. However, this latter factor is not directly
observable, again, especially for proprietary \hbox{systems}. However, we {\em
can} measure the number of plans actually generated by the optimizer
when presented with different cardinalities of the underlying tables.
We term this set of plans the ``effective plan
space'' and term the number of such plans, the cardinality of the
effective plan space, or ``{\em CEPS}''. Similarly, we cannot directly observe
the cost model utilized by the optimizer for the operators it considers, but
can classify the cost model of each operator as continuous (smoothly varying
with cardinality of its input(s) and without jumps) or discontinuous
(sometimes producing observable jumps in the predicted
cost), and can thus count ``the number of operators with
discontinuity.'' (Such operators have also been termed
``nonlinear''~\cite{Hulgen03}, though we focus on a more specific aspect of
discontinuity.) Finally, the variable ``Number of repeats'' is the number
of times a plan is reused across the cardinality of the variable table. For
example, if the sequence of plans for a query as the cardinality increases
is $A ~ B ~ B ~ B ~ C ~ C ~ A ~ B ~ B~ B ~ C ~ B ~ C ~ C ~ C ~ B$, then the
``Number of repeats'' will be six: removing sequential duplicates results in $A ~ B ~ C
~ A ~ B ~ C ~ B ~ C ~ B$, with the last six distinct plans being duplicates
of the first three. Interestingly, CEPS plus the number of repeats gives you the
number of plans in the sequence, again, after removing sequential
duplicates, nine plans in this particular case.

Plan space complexity is an {\em intervening} variable, in that it is dependent
on some of the variables on its left but is observable and thus  exerts influence on the dependent
variable on its right. We can observe these variables within experiments and
indirectly influence their value but
cannot directly intervene to specify their value. For example, we can influence the CEPS through
manipulating the values for the independent variables of optimizer and query
complexity, but cannot directly set a value for CEPS within an
experiment.

@@

The model also has two interactions within plan space
complexity, between 
CEPS and number of operators with discontinuity variables and between number of repeats and CEPS.

We hypothesize that greater plan space complexity should make it more
difficult to optimize the query.  It is important to note that the
occurrences of sub-optimality is not necessarily dependent on the presence
of a discontinuous plan operator. However, we predict that when
discontinuous plan operators appear in candidate plans, they may introduce
complexity to query optimization, especially at the cardinality where
discontinuity is possible. This is because the performance of such operators
is sensitive to the input size in a rather complex way. Any inaccuracy in
the estimation of plan statistics may lead the optimizer to select a suboptimal
plan.

\vspace{0.6em}\noindent
{\bf Hypothesis 4}: Plan space complexity (CEPS) and Number of operators
with discontinuity will be positively correlated.

@@

In this section we specify how we operationalized each of the seven
variables in the model.  Recall that each independent variable is a property
of a \hbox{DBMS} (number of operators available), of the schema (presence
  of secondary indexes), of a query (number of correlation names,
presence of an aggregate, presence of primary key attribute, and
presence of subquery), of the data (percentage of skewed data), of
the plan space (number of repeats, CEPS, and number of discontinuous
operators). There is 
one dependent variable of our model (suboptimality).

>>>>

\begin{abstract}
The query optimization phase within a \hbox{DBMS} ostensibly finds the fastest query
execution plan from a potentially large set of enumerated plans, all of
which correctly compute the specified query. Occasionally the optimizer
selects the wrong plan, for a variety of reasons. {\em Suboptimality} is
indicated by the existence of a query plan that performs more efficiently
than the DBMS's chosen plan, for the same query. From an engineering
perspective, it is of critical importance to understand the prevalence
of suboptimality and its causal factors.
We propose a novel structural causal model to explicate the
relationship between various factors in query optimization and
suboptimality. Our model associates suboptimality with the factors of
complexity of the query and optimizer and concomitant unanticipated
interactions among the components of the optimizer. This model induces a number of specific
hypothesis that were subsequently tested on multiple \hbox{DBMSes}. We
observe that \todo{Rick: check these}suboptimality prevalence
correlates positively with the number of operators available in the \hbox{DBMS}
(one proxy for optimizer complexity) and with the number of plans generated
by the optimizer (another proxy for optimizer complexity) and with
the number of correlation names (a measure of query complexity), providing
empirical support for this model as well as implications for fundamental
improvements of these optimizers. The results of this study also uncovered evidence of a potential fundamental
limitation of cost-based optimization.\todo{Rick:be more specific! Couple
  with contributions and conclusions}
This paper thus provides
a new methodology to study mature query optimizers, that of empirical
generalization, proposes a novel causal model for query suboptimality, and
tests several hypotheses deriving from this causal model.
\end{abstract}

% A category with the (minimum) three required fields
%\category{H.4}{Information Systems Applications}{Miscellaneous}
%A category including the fourth, optional field follows...
%\category{D.2.8}{Software Engineering}{Metrics}[complexity measures,
%performance measures]
\terms{}
\keywords{}
\acmformat{\doubleblind{Richard T.~Snodgrass, Sabah Currim, and Young-Kyoon
    Suh, 2017, A Causal Model of \hbox{DBMS} Suboptimality}{A Causal Model of
    \hbox{DBMS} Suboptimality}}

@@

\section{Introduction}\label{sec:intro}
DBMSes underlie all information systems and hence optimizing their performance is of
critical importance. The DBMS's query optimizer plays an important role. But
what if the optimizer {\em doesn't}: what if it selects the wrong plan?

This paper provides a thorough investigation into \hbox{DBMS} {\em
  suboptimality}: when the \hbox{DBMS} chooses a slower plan over a faster
plan for a given query. We systematically examine the factors influencing
the number of suboptimal queries. There could be multiple causes of the
suboptimality. One possible cause could be some peculiarity within the tens
of thousands of lines of code of that query optimizer. Another possible
cause could be the query's complexity. Prior research in other domains shows
that increasing complexity negatively influences
performance~\cite{campbell88,moody98}. A third possible reason could be some
fundamental limitation {\em within the general architecture of cost-based
  optimization} that will always render a number of queries suboptimal.

To better understand the impact of different factors on suboptimality of
query performance and the interaction between operators, especially in a
dynamic environment, an experimental approach is needed.
Our research introduces a novel approach
to better understand the factors influencing query performance in
\hbox{DBMSes}. Based on existing research and general knowledge of \hbox{DBMSes}, we
propose an innovative predictive model to better understand the presence and influence
of suboptimality in query evaluation. We use an experimental methodology on
a collection of \hbox{DBMSes} as subjects to test our hypotheses with respect to factors
influencing suboptimality, utilizing
experiment data collected over a \hbox{cumulative} 16,000 hours (over two
years) of query executions.
Our research falls within creative development of
new evaluation methods and metrics for artifacts, which were identified as
important design-science contributions~\cite{hevner04}.

@@

\item We find that for a surprisingly large portion of queries, the plan
  chosen  by the query optimizer is not the best plan, for some cardinality
  of the underlying tables.

\item We propose a {\em predictive model} for \hbox{DBMSes} to better
  understand the factors causing suboptimality.

\item We test the seven quite specific hypotheses that arise from that
  model across a wide range of DBMSes, queries, and data.

\item We show that the hypothesized interactions are \todo{Rick:check after Sabah} supported,
  lending credibility to our particular model.

\item Through an innovative
  analysis, we track the net cumulative benefit of a succession of operators
  added to the DBMS, and show that current relational DBMSes may have
  already reached the stage where it is difficult to provide performance
  improvement via new operators, and in fact performance may degrade.

\item The predictive model and these experimental results suggest several specific
  engineering directions.

\item A subsequent analysis implies that a new approach, fundamentally
  different from that utilized over the paper forty years, may be required
  to get past a fundamental limitation uncovered in this research.
\end{itemize}

@@

One might ask, shouldn't the task of optimizing queries be left to DBAs?  In
databases, and especially in data warehouses, the number of users writing
and running queries has been growing exponentially. This growth is aided, in
part, by the drag and drop query tools provided by the different
systems. For example, subject areas allow business users to write queries
without knowing anything about the underlying database structure. This,
coupled with constantly growing data presents new challenges for tuning. For
example, \doubleblind{the data warehouse deployed at the University of
  Arizona}{a large data warehouse we are familiar with} runs about
30,000 queries on a daily basis. 
Also, tuning a subject area or tables for one group of queries can
negatively impact the performance of other queries. Query optimization experts often take hours to tune and test existing canned
queries; the amount expended on this one system for manual query optimization
approaches \$100K/year. Therefore, we argue that it is important to
understand how existing query optimizers can be further improved, and
indeed, whether  fundamental limitation inherent in these optimizers exist
and whether such limitations are indeed already being encountered.

@@

We study these aspects \c2j{}{across \hbox{DBMSes}} to identify the
underlying causes. We have developed a predictive causal model that
identifies four 
factors that may play a role in suboptimality. The ultimate goal is to {\em
  understand} a component within a \hbox{DBMS}, its cost-based optimizer, through the
articulation and empirical testing of a general scientific theory.

In Section~\ref{sec:related} we briefly summarize the vast amount of related
work in query optimization to establish the technical basis for our study.
The following section introduces the methodology we will follow, that of
{\em empirical generalization}~\cite{cohenbook}. We then present in
Sections~\ref{sec:model} and~\ref{sec:operationalization} a predictive,
causal model of suboptimality and state seven specific hypotheses derived
from that model. We then test these hypotheses across a number of queries,
data, and cardinalities, whose results provide strong support for the
validity of the proposed model.  These are the first scientific results that
we are aware of that apply {\em across} \hbox{DBMSes}, rather than on a
single, specific \hbox{DBMS} or on a specific algorithm. In
Section~\ref{sec:diminishing} presents a follow-on analysis that
demonstrates a fundamental limit to the number of operators that a
cost-based optimizer can support, and indicates that perhaps extant DBMSes
have hit that limit. This model has implications listed in
Section~\ref{sec:engineering} for research in engineering more efficient
\hbox{DBMSes}.

@@

The upper graph in Figure~\ref{fig:query769} represents the plans chosen by
a common DBMS as the cardinality of {\tt FT\_HT1} decreases from 2M tuples
to 10K tuples in units of 10K tuples. The \hbox{x-axis} depicts the estimated
cardinality and the y-axis  a plan chosen for an interval of
cardinalities. So Plan~P0 was chosen for 2M tuples, switching to Plan~P1 at
1,830,000 tuples, back to Plan~P0 at 1,640,000 tuples, and so on, through
the plan
sequence P0, P1, P0, P1, P2, P0, P2, and finally P3 at 40,000 tuples.

@@

We have found through our experiments that flutter and suboptimality are all
around us: {\em every} \hbox{DBMS} that we have examined, including several
open source \hbox{DBMSes} and several proprietary
\hbox{DBMSes}, covering much of the installed base worldwide, exhibit these
phenomena, even when optimizing very simple queries. In the Confirmatory
Experiment described in detail in Section~\ref{sec:experiments}, we started
with 6,967 query instances (a query run on a specific \hbox{DBMS}) after \doubleblind{our}{an} extensive query measurement
protocol~\cite{TTPv1,TTPv2} applied its extensive sanity checks. While about
20\%, or 1,491, of these query instances contained only one
plan, a few of the other query instances switched plans at almost every change in cardinality (we varied the
cardinality in increments of 10K tuples, a total of 200 cardinalities): see
Figure~\ref{fig:planchanges}. Slightly over half, or 3,933 query instances, exhibited
suboptimality somewhere along those 200 cardinalities; a few had
many changes to a plan that was in fact suboptimal, as indicated in
Figure~\ref{fig:suboptplanchanges}, across all four \hbox{DBMSes} considered.

\begin{figure}[h!]\centering
\caption{Cumulative percentage of queries \hbox{exhibiting} the \hbox{indicated}
  number of plan changes\label{fig:planchanges}}
\end{figure}

\begin{figure}[t]\centering
\includegraphics[width=0.60\textwidth]{figures/zi_ncpq.pdf}
\caption{Cumulative percentage of queries exhibiting the \hbox{indicated}
  number of changes to a {\em suboptimal} plan\label{fig:suboptplanchanges}}
\vspace*{-2ex}
\end{figure}

One oft-stated observation is that the role of query optimization is not to
get the {\em best} plan, but rather to get a plan that is acceptably
good. (Thus, the very term ``query optimizer'' is aspirational rather than
accurate.) Figure~\ref{fig:suboptcumulative} shows
the cumulative distribution of the relative amount of suboptimality (where
an $x$-value of 100 denotes that the query ran 100\% slower than the optimal
query, that is, twice as long). The good news is that 2,738 query instances,
or 67\% of the suboptimal queries, exhibited only a small degree of
suboptimality: less than 30\%. The challenge is that over fifth of all
queries (1,355) exhibited a significant amount of suboptimality ($\geq
30\%$). The relative slowdown can thus be quite large for some queries.

We started with 7,640 query instances, each of which is a particular query running on a
particular DBMS, cf.~Experiment 7 of Table~\ref{tab:run_stat}
in Section~\ref{sec:experiments}. After our protocol, we were left with
  6,967 query instances, of which 5,475 had at least one change point, so those are
  the ones we consider further. Of those, 1,382 had {\em no} suboptimality, so
4,093 (75\%) had some suboptimality.

@@

We emphasize four important points.
\begin{itemize}
\item First, we used a sophisticated query
measurement methodology that reduces the measurement variance, so that the
query plans we identify as suboptimal definitely are so.
\item Second, these
results are over four \hbox{DBMSes}, and thus, such phenomena are not dependent on
a particular implementation of cost-based optimization. Rather, they seem to
be common to {\em any} cost-based optimizer, independent of the specific
cardinality \hbox{estimation} or plan costing or plan enumeration algorithm or
implementation.
\item Third, suboptimal plans are {\em not} the result of poor
coding or of inadequate algorithms. We view query optimization in modern
\hbox{DBMSes} as an engineering marvel, especially given the complexity of the SQL
language and the requirements and expectations of \hbox{DBMS} users, who often
demand that important queries simply not get slower with a new release of
the \hbox{DBMS}. Rather, the prevalence of suboptimality observed here is a
reflection of the complexity of the task of query optimization.
\item Fourth, we wanted to understand whether a fundamental
limitation exists in the prevalent approach to query optimization utilized
by DBMSes generally, and certainly by the four DBMSes that we studies.
\end{itemize}

This is why we needed a new methodology. We want to understand
cost-based optimization deeply. This means that we need to go beyond the
examination of a single query optimizer, as is done in almost every paper
over the forty-year history of query optimization research, to study multiple instances of
that optimization architecture, in an effort to achieve generalizable
results.

Thus, in this paper,
we articulate a predictive causal model for how
and in what circumstances suboptimality arises and provide compelling
evidence via \hbox{hypothesis} testing that this model accurately characterizes the
\hbox{behavior} of query optimizers in general. This is what is meant by ``empirical
generalization''~\cite{cohenbook} and why it is needed to answer such questions. In
  the long history of research in database query optimization, or even of
  databases in general, our model and its hypothesis tests are the first
  predictive results that we are aware of that apply {\em across}
  \hbox{DBMSes}, rather than on a single, specific \hbox{DBMS} or on a
  specific algorithm. Such a DBMS-agnostic, though paradigmatic, causal model can
\hbox{provide} guidance to the community about where fundamental research is needed
and to the \hbox{DBMS} engineers about where to focus their efforts.

@@

\todo{Rick: For Science article, add history: IMS->System R,
  procedural->declarative, Selinger's query opt algorithm enabling rise of
  relational model}
\todo{Rick: For TODS, shorten history, probably remove SQL cites, just one
  sentence about central role of Selinger}
SQL has emerged as the {\em de facto} and {\em de jure} standard relational
database query and update language. It emerged as an ANSI and ISO standard
in 1986--7, with a major extension and refinement as SQL-92~\cite{Melton93} 
and an even larger extension as SQL:1999~\cite{Melton03} and refinement as
SQL:2008~\cite{SQL2008} and SQL:2011~\cite{Melton11}.

There has been extensive work in query optimization over the last 40
years~\cite{Ioannidis96,Jarke84}, during which a particular quite effective
paradigm had taken hold, in both open-source and proprietary DBMSes. In this
over-arching paradigm, query optimization and evaluation
proceeds in several general steps~\cite{Ramakrishnan03}. First, the SQL query is 
translated into alternative query evaluation plans based on the 
relational algebra via {\em query enumeration}. The cost
of each plan is then estimated and the plan with the lowest estimated cost
is chosen. These steps comprise query optimization, specifically {\em
cost-based query optimization}~\cite{Selinger}. The selected query plan is
then evaluated by the query execution engine which implements a set of
physical operators, often several for each logical operator such as
join~\cite{Graefe93}.

@@

An optimizer for a language like SQL must contend with a huge search space of
complex queries. Its first objective must be {\em correctness}: that the
resulting query evaluation plan produce the correct result for the
query. This objective must be ensured both by the initial SQL-to-relational
algebra translator and by the subsequent query enumerator. A secondary but clearly very important objective is {\em efficiency}; after
all, that is the raison d'etre for this phase. As is well known and has been
alluded to already, the name
for this phase is an exaggeration, as existing optimizers do not produce
provably optimal plans. That said, the
query optimizers of prominent \hbox{DBMSes} generally do a superb job of producing
the best query evaluation plan for most queries. This performance is the
result of a fruitful collaboration between the research community and
developers.

@@

However, even with great effort over decades, optimizers as a general class
are still poorly understood. As has been observed, ``query optimization has
acquired the dubious reputation of being something of a black
art''~\cite{Babcock05}. DeWitt has gone farther, stating that ``query
optimizers [do] a terrible job of producing reliable, good plans [for
  complex queries] without a lot of hand
tuning''~\cite[page~59]{winslett02}. And as we will see, suboptimality may
occur in sophisticated query optimizers even when considering only simple queries.

While this paper does not provide direct solutions to address
suboptimality, we envision that by following up on the implications of our proposed predictive model
for suboptimality, engineering practice, such as dynamic reoptimization just
mentioned, may benefit. We elaborate on this subject in
Sections~\ref{sec:root} and \ref{sec:engineering}, where we discuss the engineering implications
of our causal model.

@@

For the construct of {\em query complexity} we have identified four
variables: ``Number of correlation names'', ``Presence of aggregate'',
``Presence of primary key attribute'', and ``Presence of subquery''. For each independent
variable, we have interventional control in our experiments, in that we can
manipulate the values of these variables through the construction of the
actual query to be optimized. The first is the number of correlation names
defined in the FROM clauses. The second is whether a single aggregate operator
(that is, either {\tt Max} or {\tt Avg}) appears in the SELECT clause.
The third
is whether a primary key attribute appears in at least one predicate in the
WHERE clause. The fourth is whether a single subquery appears in the
WHERE clause; that subquery evaluates to a single value that is
equality-compared with an attribute, in the where clause. The rationale is
that each factor may expand the number of plans or otherwise render the search for
the optimal plan more complex.

@@

In the middle of the figure is the construct of {\em plan space
complexity}. Given a particular query, its
complexity will impact the total number of candidate plans considered
by the optimizer. This variable is not directly
observable, again, especially for proprietary \hbox{systems}. However, we {\em
can} measure the number of plans actually generated by the optimizer
when presented with different cardinalities of the underlying tables.
We term this set of plans the ``effective plan
space'' and term the number of such plans, the cardinality of the
effective plan space, or ``{\em CEPS}''. Similarly, we cannot directly observe
the cost model utilized by the optimizer for the operators it considers, but
can classify the cost model of each operator as continuous (smoothly varying
with cardinality of its input(s) and without jumps) or discontinuous
(sometimes producing observable jumps in the predicted
cost), and can thus count ``the number of operators with
discontinuity.'' (Such operators have also been termed
``nonlinear''~\cite{Hulgen03}, though we focus on a more specific aspect of
discontinuity.) Finally, the variable ``Number of repeats'' {\em is}
directly measurable as the number
of times a plan is reused across the cardinality of the variable table. For
example, if the sequence of plans for a query as the cardinality increases
is $A ~ B ~ B ~ B ~ C ~ C ~ A ~ B ~ B~ B ~ C ~ B ~ C ~ C ~ C ~ B$, then the
``Number of repeats'' will be six: removing sequential duplicates results in $A ~ B ~ C
~ A ~ B ~ C ~ B ~ C ~ B$, with the last six distinct plans being duplicates
of the first three. Interestingly, CEPS plus the number of repeats gives you the
number of plans in the sequence, again, after removing sequential
duplicates, nine plans in this particular case. Thus, we associate with the latent construct
of plan space complexity three measurable (that is, indirect) variables. Our
  model stipulates that the construct, and thus the three associated with
  this construct, intervene between the constructs of optimizer complexity,
  schema complexity and query complexity and the construct of suboptimality.

Plan space complexity is an {\em intervening} construct, in that it is dependent
on some of the constructs on its left but is observable and thus exerts influence on the dependent
variable on its right. We can observe these variables within experiments and
indirectly influence their value but
cannot directly intervene to specify their value. For example, we can influence the CEPS through
manipulating the values for the independent variables of optimizer and query
complexity, but cannot directly set a value for CEPS within an
experiment.

@@

The model also has two interactions within plan space
complexity, between 
CEPS and number of operators with discontinuity variables and between number of repeats.

We hypothesize that greater plan space complexity should make it more
difficult to optimize the query.  It is important to note that the
occurrences of suboptimality is not necessarily dependent on the presence
of a discontinuous plan operator. However, we predict that when
discontinuous plan operators appear in candidate plans, they may introduce
complexity to query optimization, especially at the cardinality where
discontinuity is possible. This is because the performance of such operators
is sensitive to the input size in a rather complex way. Any inaccuracy in
the estimation of plan statistics may lead the optimizer to select a suboptimal
plan.

\vspace{0.6em}\noindent
{\bf Hypothesis 4}: Plan space complexity (CEPS) and (a)~Number of operators
with discontinuity and (b) Number of repeats will be positively correlated.

@@

In this section we specify how we operationalized each of the seven
variables in the model.  Recall that each independent variable is a property
of a \hbox{DBMS} (number of operators available), of the schema (presence
  of secondary indexes), of a query (number of correlation names,
presence of an aggregate, presence of primary key attribute, and
presence of subquery), of the data (percentage of skewed data), and of
the plan space (number of repeats, CEPS, and number of discontinuous
operators). There is 
one dependent variable of our model (suboptimality).
